Ethereum: A Secure Decentralised Generalised Transaction Ledger

### 2014-04-19 Paper.tex
Trie & serialisation format folded into main definition.

Scalars and fixed-size byte vectors are denoted with a normal lower-case letter, \eg $n$ is used in the document to denote a transaction nonce. Those with a particularly special meaning may be greek, \eg $\delta$, the number of items required on the stack for a given operation.

@@

@@

@@

The world state (\textit{state}, see Appendix \ref{app:state}), is a mapping between addresses (160-bit identifiers) and account states (a data structure serialised as RLP, see Appendix \ref{app:rlp}). Though not stored on the blockchain, it is assumed that the implementation will maintain this mapping in a modified Merkle Patricia tree (\textit{trie}, see Appendix \ref{app:trie}). The trie requires a simple database backend that maintains a mapping of bytearrays to bytearrays; we name this underlying database the state database. This has a number of benefits; firstly the root node of this structure is cryptographically dependent on all internal data and as such its hash can be used as a secure identity for the entire system state. Secondly, being an immutable data structure, it allows any previous state (whose root hash is known) to be recalled by simply altering the root hash accordingly. Since we store all such root hashes in the blockchain, we are able to trivially revert to old states.

@@

\begin{description}
\item[nonce] A scalar value equal to the number of transactions sent from this address, or, in the case of contract accounts, the number of contract-creations made by this account. For account of address $a$ in state $\mathbb{S}$, this would be formally denoted $\mathbb{S}[a]_n$.
\item[balance] A scalar value equal to the number of wei owned by this address. Formally denoted $\mathbb{S}[a]_b$.
\item[stateRoot] A 256-bit hash equal to the root node of a further trie structure that encodes the storage contents of the contract. Though a separate data structure, this trie is still stored using the same underlying state database. This trie takes the form as a simple mapping between domains of 256-bit values. The storage, modelled as a $2^{256}$ size sequence of 256-bit values is formally denoted $\mathbb{S}[a]_\mathbf{s}$.
\item[codeHash] The hash of the EVM code of this contract---this is the code that gets executed should this address receive a message call; it is immutable and thus, unlike all other fields, cannot be changed after construction. All such code fragments are contained in the state database under their corresponding hashes for later retrieval. This hash is formally denoted $\mathbb{S}[a]_c$, and thus the code is denoted as $b$, given that $\texttt{\small SHA3}(b) = \mathbb{S}[a]_c$.
\end{description}

@@

@@

\begin{description}
\item[nonce] A scalar value equal to the number of transactions sent by the sender; formally $T_n$.
\item[value] A scalar value equal to the number of wei to be transferred to the message call's recipient or, in the case of contract creation, as an endowment to the newly created contract's account; formally $T_v$.
\item[gasPrice] A scalar value equal to the number of wei to be paid per unit of gas for all computation costs incurred as a result of the execution of this transaction; formally $T_p$.
\item[gasLimit] A scalar value equal to the maximum amount of gas that should be used in executing this transaction. This is paid up-front, before any computation is done and may not be topped-up later; formally $T_g$.
\item[to] The 160-bit address of the message call's recipient or the zero address for a contract-creation transaction; formally $T_r$.
\item[v, r, s] A scalar and complex value corresponding to the signature of the transaction and used to determine the sender of the transaction; formally $T_v$ and $T_{rs}$. This is expanded in Appendix \ref{app:signing}.
\end{description}

@@

@@

\begin{description}
\item[parentHash] The SHA3 256-bit hash of the parent block, in its entirety; formally $H_p$.
\item[unclesHash] The SHA3 256-bit hash of the uncles list portion of this block; formally $H_u$.
\item[coinbase] The 160-bit address to which all fees collected from the successful mining of this block be transferred; formally $H_b$.
\item[stateRoot] The SHA3 256-bit hash of the root node of the state trie, after all transactions are executed and finalisations applied; formally $H_r$.
\item[transactionsHash] The SHA3 256-bit hash of the transactions list portion of this block; formally $H_t$.
\item[difficulty] A scalar value corresonding to the difficulty level of this block. This can be calculated from the previous block's difficulty level and the timestamp; formally $H_d$.
\item[timestamp] A scalar value equal to the reasonable output of Unix's time() at this block's inception; formally $H_s$.
\item[extraData] An arbitrary byte array containing data relavant to this block. With the exception of the genesis block, this must be 32 bytes or fewer. Not formally used.
\item[nonce] A 256-bit hash which proves that a sufficient amount of computation has been carried out on this block; formally $H_n$.
\end{description}

The other two components in the block are simply a list of uncle block headers (of the same format as above) and a list of the transactions (as defined previously). Formally, we can refer to a block $B$:
\begin{equation}
B \equiv (H, \mathbf{T}, \mathbf{U})
\end{equation}

We can assert its validity if and only if it satisfies several conditions: It must be internally consistent with the uncle and transaction block hashes and the given transactions $\mathbf{T}$, when executed in order on the base state $\mathbb{S}$, result in a new state of the identity $H_r$:
\begin{eqnarray}
& & \mathtt{\tiny SHA3}(\mathtt{\tiny RLP}(\mathbf{U})) \equiv H_u \\
& \wedge & \mathtt{\tiny SHA3}(\mathtt{\tiny RLP}(\mathbf{T})) \equiv H_t \\
& \wedge & r(\Pi(\mathbb{S}, B)) \equiv H_r
\end{eqnarray}

where $r(\mathbb{S})$ is the root node hash of the Merkle Patricia tree structure containing the key-value pairs of $\mathbb{S}$ with values encoded using RLP.

@@

We define intrinsic gas $g_0$, the amount of gas this transaction requires to be paid prior to execution, as follows:
\begin{equation}
g_0 \equiv \begin{cases}
g_d + G_{create} & \text{if} \quad T_r = 0 \\
g_d + G_{call} & \text{otherwise} \\
\end{cases}
\end{equation}

@@

\begin{equation}
g_d \equiv \begin{cases} (\lVert T_\mathbf{i}\rVert + \lVert T_\mathbf{b}\rVert) G_{data} & \text{if} \quad T_r = 0 \\
 \lVert T_\mathbf{d}\rVert G_{data} & \text{otherwise}
 \end{cases}
\end{equation}

@@

Evaluating $\mathbb{S}_P$ from $\mathbb{S}_0$ depends on the transaction type; either contract creation or message call; we define the pairing of post-execution provisional state ($\mathbb{S}_P$) and remaining gas ($g'$):
\begin{equation}
(\mathbb{S}_P, g') \equiv \begin{cases}
\Lambda(\mathbb{S}_0, S(T), T_o, g, T_p, T_v, T_\mathbf{i}, T_\mathbf{b}) & \text{if} \quad T_r = 0 \\
\Theta^*(\mathbb{S}_0, S(T), T_o, T_r, g, T_p, T_v, T_\mathbf{d}) & \text{otherwise}
\end{cases}
\end{equation}

@@

The exceptional halting function $Z$ is defined as:
\begin{equation}
Z(\mathbb{S}, \mathbb{M}, I) \equiv \mathbb{M}_g < C(\mathbb{S}, \mathbb{M}, I) \vee \mathbf{\delta}_w = \varnothing \vee \lVert\mathbb{M}_\mathbf{s}\rVert < \mathbf{\delta}_w
\end{equation}

@@

If RLP is used to encode a positive integer, it must be specified as the shortest byte array such that the big-endian interpretation of it is equal. Thus the RLP of some positive integer $i$ is defined as:
\begin{equation}
\mathtt{\tiny RLP}(i : i \in \mathbb{P}) \equiv \mathtt{\tiny RLP}(\mathtt{\tiny BE}(i))
\end{equation}

@@

A branch is then only used when necessary; no branch nodes may exist that contain only a single non-zero entry. We may formally define this structure with the structural composition function $c$:
\begin{equation}
c(\mathfrak{I}, i) \equiv \begin{cases}
\texttt{\small RLP}\Big( \big(\texttt{\small HP}(I_0[i .. (\lVert I_0\rVert - 1)], I_1 \big) \Big) & \text{if} \quad \lVert \mathfrak{I} \rVert = 1 \quad \text{where} \; \exists I: I \in \mathfrak{I} \\
\texttt{\small RLP}\Big(\big(\texttt{\small HP}(I_0[i .. (j - 1)]), n(\mathfrak{I}, j) \big) \Big) & \text{if} \quad i \ne j \quad \text{where} \; j = \arg \max_x : \exists \mathbf{l}: \lVert \mathbf{l} \rVert = x : \forall_{I \in \mathfrak{I}}: I_0[0 .. (x - 1)] = \mathbf{l} \\
\texttt{\small RLP}\Big( (u(0), u(1), ..., u(15), v) \Big) & \text{otherwise} \quad \text{where} \begin{array}[t]{rcl}
u(n) & \equiv & n(\{ I : I \in \mathfrak{I} \wedge I_0[i] = n \}, i + 1) \\
v & = & \begin{cases}
I_1 & \text{if} \quad \exists I: I \in \mathfrak{I} \wedge \lVert I_0 \rVert = i \\
() & \text{otherwise}
\end{cases}
\end{array}
\end{cases}
\end{equation}

@@

\section{Formal Specification of Structures}
Though RLP is data-agnostic, it does specify a canonical representation for integer quantities. It is big-endian with no leading zero-bytes. Thus for elements than feasibly can be stored as integers, it becomes important to specify whether they should adhere to this canonical representation or be left in some other (perhaps more 'native') format.

In the case of counts, balances, fees and amounts of wei, the canon-integer form must be used when storing in RLP. We call these INTs.

In the case of hashes (256-bit or 160-bit), user-visible strings and specialised byte-arrays (e.g. hex-prefix notation from the trie), they should be stored as unformatted byte-array data and not altered into some other form. We call these BINs.

When interpreting RLP data, clients are required to consider non-canonical INT fields in the same way as otherwise invalid RLP data and dismiss it completely.

Specifically:

for the Block header:
\begin{verbatim}
[
  parentHash: BIN (256-bit),
  unclesHash: BIN (256-bit),
  coinbase: BIN (160-bit),
  stateRoot: BIN (256-bit),
  transactionsHash: BIN (256-bit),
  difficulty: INT,
  timestamp: INT,
  extraData: BIN (typically 256-bit),
  nonce: BIN (256-bit)
]
\end{verbatim}

(note: 'nonce', the last element, refers to a hash here and so is binary)

for entries in the State trie for normal addresses:
\begin{verbatim}
[
  balance: INT,
  nonce: INT
]
\end{verbatim}

and for contract addresses:
\begin{verbatim}
[
  balance: INT,
  nonce: INT,
  storageRoot: BIN (256-bit),
  codeHash: BIN (256-bit)
]
\end{verbatim}

(note: 'nonce', the second element, refers to a tx-count here and so is integer)

for message call transactions:

\begin{verbatim}
[
  nonce: INT,
  value: INT,
  gasPrice: INT,
  gas: INT,
  recvAddr: BIN (160-bit),
  data: BIN (arbitrary length),
  v: INT,
  r: INT,
  s: INT
]
\end{verbatim}

or, for contract creation transactions,

\begin{verbatim}
[
  nonce: INT,
  value: INT,
  gasPrice: INT,
  gas: INT,
  0: BIN (160-bit),
  code: BIN (arbitrary length),
  init: BIN (arbitrary length),
  v: INT,
  r: INT,
  s: INT
]
\end{verbatim}

(note: 'nonce', the first element, refers to a tx-count here and so is integer)

The nonce in the transaction refers to the total amount of transactions sent from the address up until that moment in time.

for blocks, there are no immediate data field, but lists:

\begin{verbatim}
[
  blockHeader: [...]
  txList: [ [...], [...], ... ]
  uncleList: [ [...], [...], ... ]
]
\end{verbatim}

Uncle-blocks contain only the uncle's header.

>>>>

Scalars and fixed-size byte sequences (or, synonymously, arrays) are denoted with a normal lower-case letter, \eg $n$ is used in the document to denote a transaction nonce. Those with a particularly special meaning may be greek, \eg $\delta$, the number of items required on the stack for a given operation.

@@

Throughout, we assume scalars are positive integers and thus belong to the set $\mathbb{P}$. The set of all byte sequences is $\mathbb{B}$, formally defined in Appendix \ref{app:rlp}. If such a set of sequences is restricted to those of a particular length, it is denoted with a subscript, thus the set of all byte sequences of length $32$ is named $\mathbb{B}_{32}$. This is formally defined in section \ref{ch:block}.

@@

When considering the use of existing functions, given a function $f$, the function $f^*$ denotes a similar, element-wise version of the function mapping instead between sequences. It is formally defined in chapter \ref{ch:block}.

@@

The world state (\textit{state}), is a mapping between addresses (160-bit identifiers) and account states (a data structure serialised as RLP, see Appendix \ref{app:rlp}). Though not stored on the blockchain, it is assumed that the implementation will maintain this mapping in a modified Merkle Patricia tree (\textit{trie}, see Appendix \ref{app:trie}). The trie requires a simple database backend that maintains a mapping of bytearrays to bytearrays; we name this underlying database the state database. This has a number of benefits; firstly the root node of this structure is cryptographically dependent on all internal data and as such its hash can be used as a secure identity for the entire system state. Secondly, being an immutable data structure, it allows any previous state (whose root hash is known) to be recalled by simply altering the root hash accordingly. Since we store all such root hashes in the blockchain, we are able to trivially revert to old states.

@@

\begin{description}
\item[nonce] A scalar value equal to the number of transactions sent from this address, or, in the case of contract accounts, the number of contract-creations made by this account. For account of address $a$ in state $\mathbb{S}$, this would be formally denoted $\mathbb{S}[a]_n$.
\item[balance] A scalar value equal to the number of Wei owned by this address. Formally denoted $\mathbb{S}[a]_b$.
\item[stateRoot] A 256-bit hash of the root node of a trie structure that encodes the storage contents of the contract, itself a simple mapping from byte arrays of size 32 to arbitrary length byte arrays. The hash is formally denoted $\mathbb{S}[a]_s$. Since I typically wish to refer not to the trie's root hash but to the underlying set of key/value pairs stored within, I define a convenient equivalence $\texttt{\small TRIE}(\mathbb{S}[a]_\mathbf{s}) \equiv \mathbb{S}[a]_s$. It shall be understood that $\mathbb{S}[a]_\mathbf{s}$ is not a `physical' member of the account and does not contribute to its later serialisation.
\item[codeHash] The hash of the EVM code of this contract---this is the code that gets executed should this address receive a message call; it is immutable and thus, unlike all other fields, cannot be changed after construction. All such code fragments are contained in the state database under their corresponding hashes for later retrieval. This hash is formally denoted $\mathbb{S}[a]_c$, and thus the code may be denoted as $\mathbf{b}$, given that $\texttt{\small SHA3}(\mathbf{b}) = \mathbb{S}[a]_c$.
\end{description}

@@

Thus we may define a world-state collapse function $L_S$:
\begin{eqnarray}
L_S(\mathbb{S}) & \equiv & \{ p(a): \mathbb{S}[a] \neq \varnothing \} \\
p(a) & \equiv & \big(a, \texttt{\small RLP}\big( (\mathbb{S}[a]_n, \mathbb{S}[a]_b, \mathbb{S}[a]_s, \mathbb{S}[a]_c) \big) \big)
\end{eqnarray}

This function, $L$, is used alongside the trie function to provide a short identity (hash) of the world state. We assume:
\begin{equation}
\forall a: \mathbb{S}[a] = \varnothing \vee (a \in \mathbb{B}_{20} \wedge v(\mathbb{S}[a]))
\end{equation}

Where $v$ is the account validity function:
\begin{eqnarray}
v(x) & \equiv & x_n \in \mathbb{P} \wedge x_p \in \mathbb{P} \wedge (x_r = x_c = \varnothing \vee w(x)) \\
w(x) & \equiv & x_r \in \mathbb{B}_{32} \wedge \quad x_c \in \mathbb{B}_{32}
\end{eqnarray}

@@

\begin{description}
\item[nonce] A scalar value equal to the number of transactions sent by the sender; formally $T_n$.
\item[value] A scalar value equal to the number of wei to be transferred to the message call's recipient or, in the case of contract creation, as an endowment to the newly created contract's account; formally $T_v$.
\item[gasPrice] A scalar value equal to the number of wei to be paid per unit of gas for all computation costs incurred as a result of the execution of this transaction; formally $T_p$.
\item[gasLimit] A scalar value equal to the maximum amount of gas that should be used in executing this transaction. This is paid up-front, before any computation is done and may not be topped-up later; formally $T_g$.
\item[to] The 160-bit address of the message call's recipient or the zero address for a contract-creation transaction; formally $T_t$.
\item[v, r, s] A scalar and complex value corresponding to the signature of the transaction and used to determine the sender of the transaction; formally $T_w$, $T_r$ and $T_s$. This is expanded in Appendix \ref{app:signing}.
\end{description}

@@

The function $L_T$ prepares a transaction for being transformed into an RLP-serialised byte array:
\begin{equation}
L_T(T) \equiv \begin{cases}
(T_n, T_v, T_p, T_g, T_t, T_\mathbf{b}, T_\mathbf{i}, T_w, T_r, T_s) & \text{if} \; T_t = 0\\
(T_n, T_v, T_p, T_g, T_t, T_\mathbf{d}, T_w, T_r, T_s) & \text{otherwise} 
\end{cases}
\end{equation}

Here, we assume all components are interpreted by the RLP as integer values, with the exception of the arbitrary length byte arrays $T_\mathbf{b}$, $T_\mathbf{i}$ and $T_\mathbf{d}$ and the address hash $T_\mathbf{t}$:
\begin{equation}
\begin{array}{lclclcl}
T_n \in \mathbb{P} & \wedge & T_v \in \mathbb{P} & \wedge & T_p \in \mathbb{P} & \wedge & T_g \in \mathbb{P} \\
& \wedge & T_w \in \mathbb{P} & \wedge & T_r \in \mathbb{P} & \wedge & T_s \in \mathbb{P} \\
& \wedge & T_\mathbf{d} \in \mathbb{B} & \wedge & T_\mathbf{b} \in \mathbb{B} & \wedge & T_\mathbf{i} \in \mathbb{B} \\
& \wedge & T_t \in \mathbb{B}_{20} &&\\
\end{array}
\end{equation}

@@

\begin{description}
\item[parentHash] The SHA3 256-bit hash of the parent block, in its entirety; formally $H_p$.
\item[unclesHash] The SHA3 256-bit hash of the uncles list portion of this block; formally $H_u$.
\item[coinbase] The 160-bit address to which all fees collected from the successful mining of this block be transferred; formally $H_b$.
\item[stateRoot] The SHA3 256-bit hash of the root node of the state trie, after all transactions are executed and finalisations applied; formally $H_r$.
\item[transactionsHash] The SHA3 256-bit hash of the transactions list portion of this block; formally $H_t$.
\item[difficulty] A scalar value corresonding to the difficulty level of this block. This can be calculated from the previous block's difficulty level and the timestamp; formally $H_d$.
\item[timestamp] A scalar value equal to the reasonable output of Unix's time() at this block's inception; formally $H_s$.
\item[extraData] An arbitrary byte array containing data relavant to this block. With the exception of the genesis block, this must be 32 bytes or fewer; formally $H_x$.
\item[nonce] A 256-bit hash which proves that a sufficient amount of computation has been carried out on this block; formally $H_n$.
\end{description}

The other two components in the block are simply a list of uncle block headers (of the same format as above) and a list of the transactions (as defined previously). Formally, we can refer to a block $B$:
\begin{equation}
B \equiv (B_H, B_\mathbf{T}, B_\mathbf{U})
\end{equation}

We can assert its validity if and only if it satisfies several conditions: It must be internally consistent with the uncle and transaction block hashes and the given transactions $B_\mathbf{T}$, when executed in order on the base state $\mathbb{S}$ (derived from the final state of the parent block), result in a new state of the identity $H_r$:
\begin{eqnarray}
H_u & \equiv & \mathtt{\small SHA3}(\mathtt{\small RLP}(L_H^*(B_\mathbf{U}))) \\
\wedge \quad H_t & \equiv & \mathtt{\small SHA3}(\mathtt{\small RLP}(L_T^*(B_\mathbf{T}))) \\
\wedge \quad H_r & \equiv & \mathtt{\small TRIE}(L_S(\Pi(\mathbb{S}, B)))
\end{eqnarray}

where 
\begin{equation}
\mathtt{\small TRIE}(L_S(\mathbb{S})) = {P(B_H)_H}_r
\end{equation}

this $\texttt{\small TRIE}(L_S(\mathbb{S}))$ is the root node hash of the Merkle Patricia tree structure containing the key-value pairs of the state $\mathbb{S}$ with values encoded using RLP, and $P(B_H)$ is the parent block of $B$, defined directly.

The function $L_B$ and $L_H$ are the preparation functions for a block and block header respectively. Much like the transaction preparation function $L_T$, we assert the types and order of the structure for when the RLP transformation is required:

\begin{eqnarray}
\quad L_H(H) & \equiv & (H_p, H_u, H_b, H_r, H_t, H_d, H_s, H_x, H_n) \\
\quad L_B(B) & \equiv & \big( L_H(B_H), L_T^*(B_\mathbf{T}), L_H^*(B_\mathbf{U}) \big)
\end{eqnarray}

With $L_T^*$ and $L_H^*$ being element-wise sequence transformations, thus:
\begin{equation}
f^*\big( (x_0, x_1, ...) \big) \equiv \big( f(x_0), f(x_1), ... \big) \quad \text{for any function} \; f
\end{equation}

The component types are defined thus:
\begin{equation}
\begin{array}[t]{lclclcl}
H_p \in \mathbb{B}_{32} & \wedge & H_u \in \mathbb{B}_{32} & \wedge & H_b \in \mathbb{B}_{20} & \wedge \\
H_r \in \mathbb{B}_{32} & \wedge & H_t \in \mathbb{B}_{32} & \wedge & H_d \in \mathbb{P} & \wedge \\
H_s \in \mathbb{P} & \wedge & H_x \in \mathbb{B} & \wedge & H_b \in \mathbb{B}_{32} &&
\end{array}
\end{equation}

where
\begin{equation}
\mathbb{B}_n = \{ B: B \in \mathbb{B} \wedge \lVert B \rVert = n \}
\end{equation}

We now have a rogourous specification for the construction of a formal block structure. The RLP function $\texttt{\small RLP}$ (see Appendix \ref{app:rlp}) provides the canonical method for transforming this structure into a sequence of bytes ready for transmission over the wire or storage locally.

@@

We define intrinsic gas $g_0$, the amount of gas this transaction requires to be paid prior to execution, as follows:
\begin{equation}
g_0 \equiv \begin{cases}
g_d + G_{create} & \text{if} \quad T_t = 0 \\
g_d + G_{call} & \text{otherwise} \\
\end{cases}
\end{equation}

@@

\begin{equation}
g_d \equiv \begin{cases} (\lVert T_\mathbf{i}\rVert + \lVert T_\mathbf{b}\rVert) G_{data} & \text{if} \quad T_t = 0 \\
 \lVert T_\mathbf{d}\rVert G_{data} & \text{otherwise}
 \end{cases}
\end{equation}

@@

Evaluating $\mathbb{S}_P$ from $\mathbb{S}_0$ depends on the transaction type; either contract creation or message call; we define the pairing of post-execution provisional state ($\mathbb{S}_P$) and remaining gas ($g'$):
\begin{equation}
(\mathbb{S}_P, g') \equiv \begin{cases}
\Lambda(\mathbb{S}_0, S(T), T_o, g, T_p, T_v, T_\mathbf{i}, T_\mathbf{b}) & \text{if} \quad T_t = 0 \\
\Theta^*(\mathbb{S}_0, S(T), T_o, T_t, g, T_p, T_v, T_\mathbf{d}) & \text{otherwise}
\end{cases}
\end{equation}

@@

The exceptional halting function $Z$ is defined as:
\begin{equation}
Z(\mathbb{S}, \mathbb{M}, I) \equiv
\begin{array}[t]{rl}
& \mathbb{M}_g < C(\mathbb{S}, \mathbb{M}, I) \\
\vee & \mathbf{\delta}_w = \varnothing \\
\vee & \lVert\mathbb{M}_\mathbf{s}\rVert < \mathbf{\delta}_w
\end{array}
\end{equation}

@@

If RLP is used to encode a scalar, defined only as a positive integer, it must be specified as the shortest byte array such that the big-endian interpretation of it is equal. Thus the RLP of some positive integer $i$ is defined as:
\begin{equation}
\mathtt{\tiny RLP}(i : i \in \mathbb{P}) \equiv \mathtt{\tiny RLP}(\mathtt{\tiny BE}(i))
\end{equation}

When interpreting RLP data, if an expected fragment is decoded as a scalar and leading zeroes are found in the byte sequence, clients are required to consider it non-canonical and treat it in the same manner as otherwise invalid RLP data, dismissing it completely.

@@

A branch is then only used when necessary; no branch nodes may exist that contain only a single non-zero entry. We may formally define this structure with the structural composition function $c$:
\begin{equation}
c(\mathfrak{I}, i) \equiv \begin{cases}
\texttt{\small RLP}\Big( \big(\texttt{\small HP}(I_0[i .. (\lVert I_0\rVert - 1)], true), I_1 \big) \Big) & \text{if} \quad \lVert \mathfrak{I} \rVert = 1 \quad \text{where} \; \exists I: I \in \mathfrak{I} \\
\texttt{\small RLP}\Big( \big(\texttt{\small HP}(I_0[i .. (j - 1)], false), n(\mathfrak{I}, j) \big) \Big) & \text{if} \quad i \ne j \quad \text{where} \; j = \arg \max_x : \exists \mathbf{l}: \lVert \mathbf{l} \rVert = x : \forall_{I \in \mathfrak{I}}: I_0[0 .. (x - 1)] = \mathbf{l} \\
\texttt{\small RLP}\Big( (u(0), u(1), ..., u(15), v) \Big) & \text{otherwise} \quad \text{where} \begin{array}[t]{rcl}
u(j) & \equiv & n(\{ I : I \in \mathfrak{I} \wedge I_0[i] = j \}, i + 1) \\
v & = & \begin{cases}
I_1 & \text{if} \quad \exists I: I \in \mathfrak{I} \wedge \lVert I_0 \rVert = i \\
() & \text{otherwise}
\end{cases}
\end{array}
\end{cases}
\end{equation}

@@
